### 2014-08-14 doc/thesis/typedlua.tex
Adding text about object-oriented programming

\subsubsection{Objects and Classes}

Besides interfaces, Typed Lua will also introduce classes.
We want to offer classes as a standard way for writing code that
Lua programmers are already using, but without tying them to a
specific model.
Our classes will not include static attributes and will also not offer
privacy rules.

>>>>

\subsubsection{Object-Oriented Programming}

%Besides interfaces, Typed Lua will also introduce classes.
%We want to offer classes as a standard way for writing code that
%Lua programmers are already using, but without tying them to a
%specific model.
%Our classes will not include static attributes and will also not offer
%privacy rules.

Lua provides minimal support for object-oriented programming.
The basic mechanism is the \texttt{:} syntactic sugar for method
calls and method definitions.
The Lua interpreter translates \texttt{obj:method(args)} into
an operation that first evaluates \texttt{obj}, if it is a table
Lua checks if there is field \texttt{method} in this table,
and then calls it with the result of evaluating \texttt{obj} as
the first argument, followed by the result of evaluating the
argument list in the original expression.

Typed Lua uses table types along with the \emph{self type} to
represent objects, as we show in the following example:
\begin{verbatim}
    local Shape = { x = 0, y = 0 }
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

This example introduces some features that we must explain
before we show the type of \texttt{Shape}.
In Lua, the \texttt{:} syntactic sugar assigns a function to the
field with a first parameter named \texttt{self}, plus any other
parameters.
Typed Lua does the same, and assigns the type $\Self$ to the
implicit parameter \texttt{self}.
Typed Lua also adds the \texttt{const} annotation because the
$\Const$ qualifier is necessary for covariance among object types work.
Under this circumstances, the type of \texttt{Shape} is an object of
the following type:
\begin{align*}
\{ & ``x":\Number, ``y":\Number, \\
& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*}\}
\end{align*}

While \texttt{:} is syntactic sugar in plain Lua, Typed Lua uses it
to check method calls, binding any occurence of the $\Self$ type
in the type of the method to the receiver.
Indexing a method and not immediately calling it with the correct
receiver is a compile-time error:
\begin{verbatim}
    Shape.move(Shape, 10, 10)  -- Shape.x = 10 and Shape.y = 10
    Shape:move(5, 20)          -- Shape.x = 5 and Shape.y = 20
    local p = Shape.move       -- compile-time error
\end{verbatim}

Lua has a mechanism for self-like (or JavaScript-like) delegation of
missing fields in a table.
After the call \texttt{setmetatable(t1, \{ \string_\string_index = t2 \})},
Lua looks up in \texttt{t2} for any missing fields of \texttt{t1}.
Lua programmers often use this mechanism to simulate classes.

Typed Lua combines this mechanism along with interfaces to allow
programmers to define classes in a similar fashion to the definition
of modules, as we show in the following example:
\begin{verbatim}
    interface Shape
      x:number
      y:number
      const new:(self, number, number) -> (self)
      const move:(self, number, number) -> ()
    end
    local Shape = { x = 0, y = 0 }
    const function Shape:new (x:number, y:number)
      local s = setmetatable({}, { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    return Shape
\end{verbatim}

In Typed Lua, \texttt{setmetatable} is a strict primitive.
Typed Lua only allows calling \texttt{setmetatable} with
two operands.
The first operand can be any table \texttt{t1}, but the second operand
must be a table of the form \texttt{\{ \string_\string_index = t2 \}},
where \texttt{t2} can be any table.

The reason for being so strict with \texttt{setmetatable} is because
it is the only mechanism that Typed Lua has to create classes
as prototype objects.
In a \texttt{setmetatable} expression, if the type of \texttt{t2}
is a subtype of the type of \texttt{t1}, it changes \texttt{t1} to
have the same type of \texttt{t2}.

In our example, we use \texttt{setmetatable} to define a constructor
\texttt{new} to the class \texttt{Shape}.
The type of \texttt{new} is
$\Self \times \Number \times \Number \times \Value{*} \rightarrow \Self \times \Nil{*}$,
because we use \texttt{setmetatable} to declare \texttt{s} having
the same type of \texttt{self}.

In the same example, we also declare the interface \texttt{Shape}.
This allows us to export the type of the class \texttt{Shape},
which is the same type that is given by the variable \texttt{Shape}:
\begin{align*}
\{ & ``x":\Number, ``y":\Number,\\
& \Const \; ``new":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Self \times \Nil{*},\\
& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*}\}
\end{align*}

In Typed Lua, exporting interfaces allow us to use \texttt{require}
for creating prototype objects that work as classes,
along with a shortcut to the table type of the class.
This allows us to use this shortcut on type annotatoins,
as we show in the following example:
\begin{verbatim}
    local Shape = require "shape"
    local shape1 = Shape:new(0, 5)
    local shape2:Shape = Shape:new(10, 10)
\end{verbatim}

Typed Lua also uses \texttt{setmetatable} to simulate single inheritance,
as we show in the following example:

\begin{verbatim}
    local Shape = require "shape"
    interface Circle
      x:number
      y:number
      radius:number
      const new:(self, number, number, value) -> (self)
      const move:(self, number, number) -> ()
      const area:(self) -> (number)
    end
    local Circle = setmetatable({}, { __index = Shape })
    Circle.radius = 0
    const function Circle:new (x:number, y:number, radius:value)
      local c = setmetatable(Shape:new(x, y), { __index = self })
      c.radius = tonumber(radius) or 0
      return c
    end
    const function Circle:area ()
      return math.pi * self.radius * self.radius
    end
    return Circle
\end{verbatim}

A call to \texttt{setmetatable} where \texttt{t1} is a table
constructor assigned to a fresh local variable, and \texttt{t2}
is not \texttt{self} gives this variable an open type.
That is the reason why we can add new methods and override
existing ones in \texttt{Circle}.

In this example, we also call \texttt{setmetatable} to redefine \texttt{new}.
Notice how we can call the constructor of \texttt{Shape} inside
the overridden constructor.
A limitation of this class system is that the overridden constructor
must be a subtype of the original constructor,
so the type of \texttt{radius} has to be quite permissive.

In the same example, we also declare the interface \texttt{Circle},
but with a syntactical overhead to manually include the new and
overridden methods.
The interface \texttt{Circle} exports the following type:
\begin{align*}
\{ & ``x":\Number, ``y":\Number, ``radius":\Number,\\
& \Const \; ``new":\Self \times \Number \times \Number \times \Value \times \Value{*} \rightarrow \Self \times \Nil{*},\\
& \Const \; ``move":\Self \times \Number \times \Number \times \Value{*} \rightarrow \Nil{*},\\
& \Const \; ``area":\Self \times \Value{*} \rightarrow \Number \times \Nil{*}\}
\end{align*}

This table type is a subtype of the table type exported by the interface \texttt{Shape},
and we can use both shortcuts on type annotations, as we show in the following example:
\begin{verbatim}
    local Circle = require "circle"
    local circle1 = Circle:new(0, 5, 10)
    local circle2:Circle = Circle:new(10, 10, 10)
    local circle3:Shape = circle1
    local circle4:Shape = circle2
    print(circle2:area())                -- 314.15926535898
    print(circle3:area())                -- compile-time error
\end{verbatim}

In all examples, if we erase all type and \texttt{const} annotations,
and type declarations, they become valid Lua code, which run with the
same semantics as Typed Lua code.

The current version of Typed Lua has some limitations regarding
the use of \texttt{setmetatable}.
One limitation is that Typed Lua does not have polymorphism,
so programmers cannot hide the calls to \texttt{setmetatable} behind
nicer abstractions, as some Lua libraries do.
Another limitation is that Typed Lua does not support operator overloading,
so programmers cannot use \texttt{setmetatable} to change the behavior
of predefined operations, as some Lua libraries do.
